---
author: DTK
date: 2025-03-06
number-offset: 22
format: live-html
---

```{r include=FALSE}
library(mosaicCalc)
```

# 2025-03-06 class notes

Go over the cylinder minimum area for a given volume.

$$V(r, h) \equiv \pi r^2 h \ \ \ \text{and}\ \ \ A(r, h) \equiv 2 \pi r^2 + 2 \pi r h$$

Set the volume for the cylinder and use that to find h as a function of $r$ that is $h = V^\star / r^2$. 

Plug this into the formula for $A()$ to eliminate the $h$ dependence.

$$A(r) = 2 \pi r^2 + 2 V^\star / r$$

$A()$ has all the hallmarks of a function with an optimum. The first term grows with $r$, the second term gets smaller with $r$. 

The process of optimizing: set $\partial_r A(r) = 0$ and solve:

$\partial_r A(r) = 4 \pi r - -2 V^\star /r^2$. This gives

$2 \pi r_\star - V^\star /r^2_\star = 0$ or, $r_\star^3 = \frac {V^\star}{2\pi}$.


## Curvature at an argmax

Find the second derivative and from that the curvature.

$${\cal K}_f(x)  \equiv \frac{\left|\partial_{xx} f(x)\right|}{\ \ \ \ \left|1 + \left[\strut\partial_x f(x)\right]^2\right|^{3/2}}$$

But at the argmax, $\partial_x f(x^\star) = 0$ so the curvature is just 
$${\cal K}_f(x^\star) = \left|\strut\partial_{xx} f(x^\star)\right|$$
Draw the picture of the objective function with the little circle tucked under the maximum. 

## Newton Step


Maximize $f(x)$

Find zero of $g(x) \equiv \partial_x f(x)$.

Newton step: $x_{n+1} = x_n - g(x_n) / g'(x_n)$

Or, translated back to $f()$

Move from current point by $-\frac{\partial_x f(x_n)}{\partial_{xx}f(x_n)}$

Sort out what are units of this "move by" quantity.

Chapter 24.

[Show Nelder-Mead animation](https://www.youtube.com/watch?v=j2gcuRVbwR0) 

## Maximum likelihood

Setting: equipment failure. How long until items break? Let's give the probability of an item lasting $x$ time units.

Exponential PDF: $p(x) = \lambda e^{-x \lambda}$

Most likely to break immediately (short lifetime) but a few will have extremely long lifetime. 

```{webr-r}
slice_plot(L*exp(-x*L) ~ x, domain(x=0:20), L=1/10)
```


I measure $n$ examples, each lasts for $x_i$ time units.

The likelihood is the product of the individual PDFs. I know $x=x_i$, so the likelihood is a function of $\lambda$. 

Likelihood: $$\prod_{i=1}^n \lambda e^{-\lambda x_i}$$

Want to find $\lambda$ that maximizes this. 

```{r}
likelihood <- function(lambda, x) {
  prod(lambda * exp(-lambda*x))
}

observations <- c(3,6,2,3,9)

likelihood(.5, x = observations)

vlikelihood <- Vectorize(likelihood, "lambda")

slice_plot(
  vlikelihood(lambda, x=observations) ~ lambda, 
  domain(lambda = 0:1))
  

```

Too small. So let's look at the logarithm.

```{r}
#| warning: false
slice_plot(
  log(vlikelihood(lambda, x=observations)) ~ lambda, 
  domain(lambda = 0:1))
```

Some statistical theory. Maximum likelihood estimation.

If $x_i$ are the values of $n$ observations, then the log-likelihood function is

${\cal L}(\lambda) \equiv n \ln(\lambda) - \lambda \sum x_i$

Where is the maximum? Differentiate and solve for zero.

$$\partial_\lambda {\cal L}(\lambda) = \frac{n}{\lambda} -  \sum x_i$$
Consequently, $\lambda^* = n / \sum x_i$, the "maximimum likelihood estimator.

## Finding the variance through differentiation

Suppose we have a normal probability distribution on $x$ with variance $\sigma^2$:

$$p(x) \equiv \frac{1}{\sqrt{2 \pi \sigma^2}} \Large e^{- (x - m)^2/2\sigma^2}$$
Given this curve, I want to find $\sigma^2$. Here's one way:

$$\ln(p(x)) \equiv - \ln(2\pi) -2 \ln(\sigma) - (x-m)^2/2\sigma^2$$

Let's look at $\partial_x \ln(p(x))$:

$$\partial_x \ln(p(x)) \equiv  -2 /\sigma - 2 (x-m)/2\sigma^2$$

And now $\partial_{xx} \ln(p(x))$:

$$\partial_{xx} \ln(p(x)) \equiv  \frac{1}{\sigma^2}$$
That is, the radius at the argmax is $\sigma^2$.

So, to find the variance of a distribution, take one over the second derivative of the logarithm of the distribution at the argmax.

The likelihood is proportional to a probability distribution. So let's find its variance. Then, we can translate that into a confidence interval, a range that will include 95% of the probability.

It will be based on the curvature, more specifically 2 times the square-root of the radius of the inscribed circle. 

Here's the curvature at the argmax

$${\cal K} = \left|\partial_{\lambda\lambda} {\cal L}(\lambda)\right| = \frac{n}{\lambda^2}$$
In other words, the variance is $\lambda^2 / n$ so the 95% confidence interval is $2 \lambda / \sqrt{n}$.

We have an estimator and its variance.


