---
author: DTK
date: 2025-03-14
number-offset: 26
format: live-html
---

# 2025-03-14 class notes


[Steve Strogatz's $pi$-day article](https://www.nytimes.com/article/pi-day-math-geometry-infinity.html?smid=nytcore-ios-share&referringSource=articleShare)

Talking about convergence.

Number systems: What is 1.11111111 $\ldots$?




currency: pounds, shillings, pence

time: hours, minutes, seconds

angle, degrees, minutes seconds: show a decimal lat/long location on Google Maps

How do we know that 3.14159265358979 ... converges to a finite number. It's the sum of an infinite number of things:

$$ 3 x^0 + 1 x + 4 x^2 + 1 x^3 + 5 x^4 + 9 x^5 + 2 x^6 + \ ... \text{ at } x=1/10$$

This looks like a function. (It is a function, it's just that we are evaluating it at x=0.1 to get $\pi$)

## Convergence of an infinitely long polynomial:

$$\sum_{0}^{n\rightarrow \infty} x^{-n} = 1 + x \sum_{0}^{n \rightarrow \infty} x^{-n} $$

so

$$ \sum_{0}^{n\rightarrow \infty} x^{-n} \left(1 -x \strut\right) = 1$$

so 

$$ \sum_{0}^{n\rightarrow \infty} x^{-n} = \frac{1}{1-x} $$

Evaluate this at x = 1/10 to get the value of 1.11111 $\ldots$


Now look at $\sum_{0}^{n\rightarrow \infty} 10 x^{-n}$. It is larger than any decimal digit expansion.
## How high a power do I really need for the polynomial?

## Exponential
```{r}
myexp <- function(x) {
  if (x < 0) 1/myexp(-x)
  else {
    power <- 0
    while(x > 0.01) {
      x <- x / 2
      power <- power + 1
    }
    exp_tailor(x)^(2^power)
  }
}
exp_tailor <- function(x){
  1 + x + x^2/2 + x^3/6 #+ x^4/24
}
```

## Sinusoid

```{r}
mysin <-function(x) {
  x <- x %% (2*pi)
  if (x < 0) -mysin(-x)
  else if (x > pi) - mysin(x - pi)
  else if (x > pi/2) mysin(pi - x)
  else if (x > pi/4) cos_taylor(x - pi/2)
  else sin_taylor(x)
}

sin_taylor <- function(x) {
  x - x^3/6. + x^5/120. - x^7/5040.
}

cos_taylor <- function(x) {
  1 - x^2/2 + x^4/24 - x^6/720
}

```
