---
title: "MOSAIC Calculus Quiz 6: Prof. Kaplan"
subtitle: "April 22, 2025"
author: "Student name: __________________.            Do what you can in 20 minutes." 
format:
  typst:
    papersize: us-letter
    margin:
      left: 1cm
      right: 1cm
      top: 1cm
      bottom: 8mm
    columns: 2
    shift-heading-level-by: -1
    
crossref:
  chapters: true
  exr-title: "Question"
  exr-prefix: "Question"
---


```{r include=FALSE}
library(mosaicCalc)
knitr::opts_chunk$set(results="asis", roman=TRUE)
```{r include=FALSE}
library(devoirs)
source("../_software/software.R")
```

[**Question 7.1**]{.underline}: 

Refer to @fig-7-1. Write down the coefficients that solve for $\vec{\mathbf b}$ in terms of $\vec{\mathbf a}$ and $\vec{\mathbf c}$. 

::: {#fig-7-1}
```{r echo=FALSE}
#| warning: false 
#| out.width: 500
solve_graph(seed = 9799, integers = TRUE)
```

:::

 

 

[**Question 7.2**]{.underline}: 

Using dot products (by hand), compute numerically the projection of $\vec{\mathbf b}$ onto $\vec{\mathbf a}$ and write down numerically the two vectors that result.

$$\vec{\mathbf a} \equiv \left(\begin{array}{c} 
-4 \\ 
1 \\ 
3 \\ 
7 \\ 
5 \\ 
\end{array}\right) \ \ \ \ \text{and}\ \ \ \ \vec{\mathbf b}\equiv \left(\begin{array}{r} 
2 \\ 
1 \\ 
0 \\ 
-3 \\
2\\
\end{array}\right)$$

 

 

 

 

 

 

[**Question 7.3**]{.underline}: 

Referring to @fig-7-1, and rounding off the vector positions to integer values, 

i. draw the projection of $\vec{\mathbf b}$ onto $\vec{\mathbf c}$. 
ii. Using dot products, find the cosine of the angle between the two vectors. 
iii. Estimate the R^2^ of the projection.


 

 

 

 

[**Question 7.4**]{.underline}: 

Consider this matrix ${\mathbf M}$ and vector ${\mathbf b}$ and the task of solving ${\mathbf M}\ \vec{\mathbf x} \ = \ \vec{\mathbf b}$

 


$${\mathbf M} \equiv \left(\begin{array}{rrrr}
1 & & 3 & & 4 & & -6 \\ 
4 & & -4 & & 0 & & 8 \\ 
8 & & 0 & & 8 & & 0 \\ 
\end{array}\right)\ \ \ \ \text{and}\ \ \ \ \vec{\mathbf b}\ \equiv\ \left(\begin{array}{r} 
-5 \\ 
-2 \\ 
1 \\ 
\end{array}\right) $$

 



i. After defining `M` and `b` in R, I tried `qr.solve(M, b)`. The result was an error message, [Error in qr.solve(M, b) : singular matrix 'a' in solve`]{style = "color: blue;"}. Explain what went wrong.

 

ii. You could fix the problem by crossing out two of the vectors in ${\mathbf M}$. Figure out two that will do the job and X-them out.

 

iii. The result of the deletion in (ii) means that there will be a non-zero residual. Pencil in a new (third) vector for $\mathbf M$ that would permit *zero residual*. (Hint: don't overthink it!)

[**Note**: Flip the sheet for another question.]

[**Question 7.5**]{.underline}: 

Construct a matrix $\mathbf Q$ with mutually orthogonal vectors that spans the same space as the given $\mathbf M$. (The vectors do not need to be unit length. Let them be whatever length is easier for you.)

$${\mathbf M}\ \equiv \ \left(\begin{array}{r} 
\vec{x} & & \vec{y} & & \vec{z}\\
8 & & 1 & & 1 \\ 
4 & & 0 & & -2 \\ 
0 & & 4 & & 2\\ 
1 & & -8 & & 1 \\ 
\end{array}\right)$$

For the sake of convenience, you can refer to the columns of ${\mathbf M}$ by the names $\vec{x}, \vec{y}$ and $\vec{z}$ respectively. (Hint: It might be easier than you are thinking.)




