---
title: "MOSAIC Calculus Quiz 7: Prof. Kaplan"
subtitle: "May 2, 2025"
author: "Student name: __________________.            Do what you can in 30 minutes." 
format:
  typst:
    papersize: us-letter
    margin:
      left: 1cm
      right: 1cm
      top: 1cm
      bottom: 8mm
    columns: 2
    shift-heading-level-by: -1
    
crossref:
  chapters: true
  exr-title: "Question"
  exr-prefix: "Question"
---


```{r include=FALSE}
library(mosaicCalc)
knitr::opts_chunk$set(results="asis", roman=TRUE)
library(devoirs)
source("../_software/software.R")
make_numbering <- function(word) {
  num <- 1
  function() {
    res <- glue::glue(word)
    num <<- num + 1
    return(res)
  }
} 
Q <- make_numbering("[**Question 7.{num}**]{{.underline}}")
```




```{r}
#| echo: FALSE
triangle_wave <- function(t) {
  ifelse(t < pi/2, 
         t,
         ifelse(t > 3*pi/2,
                t - 2*pi,
                pi - t)
  )
}
# slice_plot(triangle_wave(t) ~ t, domain(t=0:(2*pi)))
Fig <- drawFpair(triangle_wave, dom = domain(t=0:(2*pi)), bottom = -pi/2)
Fig2 <-drawFpair(triangle_wave, dom = domain(t=0:(2*pi)), bottom = -pi/2, alpha = 1)
```

`r Q()` Consider this function

```{r}
#| echo: false
#| fig-align: center
#| out-width: "80%"
Fig$P1 |>
  gf_labs(y = "f(t)")
```
a. Which one of the pattern-book functions most closely resembles the $f(t)$ plotted above?

 

b. Give a formula for the anti-derivative of the function you chose to answer (a).

 

c. On the axis grid below, draw in both the function in (b) **AND** your best sketch of the anti-derivative of $f(t)$. (That is, sketch the anti-derivative of the sharp-elbowed function drawn in the previous plot as well as your function in (b).)

```{r}
#| echo: false
#| fig-align: center
#| out-width: "90%"
Fig$P2 |>
  gf_labs(y = " ") |>
  gf_lims(y = c(-2.5, 2.5))
```

d. Explain what the "constant of integration" has to do with the alignment of the two functions you sketched in (c).

 

 

 

 

 

`r Q()`

Here are some definite integrals for which, without stating anything more about the function, we give you the numerical result.

   |   |  
---------------------------|---------|------------------------------- 
$\int_{2}^{5} f(x) \,dx = 8$ | | $\int_{-5}^{-2} f(x) \,dx = 3$
   |   |  
$\int_{2}^{10} f(x) \,dx = -6$ | | $\int_{-5}^{5} f(x) \,dx = 1$
____________________|_______|____________________

Using these the facts, calculate the following integrals. If there's not sufficient information given to perform the calculation, say so.


(a) $$\int_{5}^{10} f(x) \,dx$$  | (b) $$\int_{-2}^{2} f(x) \,dx$$ 
----|-----
 | 
(c) $$\int_{-5}^{10} 3 f(x) \,dx $$ | (d) $$\int_{-5}^{5} \left| f(x) \right| \,dx$$
______________________ | _____________________
(e) $$\int_{-5}^{10} \left[3 f(x) - 2x\right] \,dx $$ | (f) $$\int_{-2}^{5} f(x)\, dt$$

 

 

 

 

 

 
 

 

 

`r Q()` Consider the function graphed below and answer as best you can.

```{r}
#| echo: FALSE
#| fig-align: center
f <- doodle_fun(~ t, seed = 223)
slice_plot(f(t) ~ t, domain(t = -5:5)) |>
  gf_theme(theme_minimal())
```

Is this function the derivative or the anti-derivative of functions?

 

`r Q()` Explain the distinctions between an "anti-derivative," a "definite," and an "indefinite" integral. 


 

 

 

`r Q()` Simplify these expressions as much as possible:

1. $$\int \frac{1}{t} dt$$

2. $$\int \frac{1}{t^2} dt$$

3. $$\int \cos(\omega t) dt$$

4.  For this one, keep in mind that $\int_{-\infty}^\infty \text{dnorm}(t) dt = 1.$
$$\int_0^\infty \text{dnorm}(t) dt$$.

5. $$\int_0^\infty \text{pnorm}(t) dt$$

 

 

 

`r Q()` A statement like this appears in the textbook. The one in the textbook is correct, but this one has a couple of typographical errors. Fix them.

$$\int_a^b f(t) dx = F(t) \left.\right|_a^b  = F(a) - F(b)$$



