---
author: DTK
date: 2025-02-25
number-offset: 18
format: live-html
---

```{r include=FALSE}
library(mosaicCalc)
library(audio)
```

# 2025-02-25 class notes

Reminder: QUIZ at end of hour, mainly on the derivatives of the pattern-book functions.

----------

## The *finite-difference* derivative operator, ${\cal D}_x$

- Takes a function as input and produces a function as output.
- An infinity of ways to define it, all variations on the same thing.

Right sided $${\cal D}_t f(t) \equiv \frac{f(t + h) - f(t)}{h}$$

Left sided: $${\cal D}_t f(t) \equiv \frac{f(t) - f(t-h)}{h}$$    
    
Centered: $${\cal D}_t f(t) \equiv \frac{f(t+h/2) - f(t-h/2)}{h}$$      
    
Centered: $${\cal D}_t f(t) \equiv \frac{f(t+h) - f(t-h)}{2 h}$$      

- Which of these is "right"? Insofar as they give the same result, they are all right. 

- Look at the value of the derivative at any specific input as a function of $h$. For small enough $h$, all these head in the same direction.

- Definition of small enough in mathematics: $\lim h \rightarrow 0$.

- Definition of small enough in numerics: Check at $h$ and at $h/8$: do these give *equivalent* results. If so, we're done.

NOTE: "equivalent" requires that we accept the idea of "close enough for our purposes."

## Higher order derivatives

Go to the board and find formulas for the 2nd- and 3rd-order derivatives of a function with one input.

To find a finite-difference formula for a second derivative, $\cal D_{t\, t} f(t)$ apply the finite-difference formula to $\cal D_t f(t)$:

$$\cal D_{t\ t} f(t) = \frac{\cal D_t f(t + h) - \cal D_t f(t-h)}{2h} = $$

$$= \frac{\frac{f(t + h + h) - f(t)}{2 h} - \frac{f(t) - f(t-h - h)}{2 h}}{2 h}$$

$$ \frac{f(t + 2 h) - 2 f(t) + f(t-2h)}{4 h^2}$$

For simplicity, let's rewrite this using a new $h_\star = 2 h$, getting
$$\cal D_{t\ t} f(t) = \frac{f(t + h_\star) - 2 f(t) + f(t - h_\star)}{2h_\star^2}$$

Paint a geometric picture of this: The average of the two end points minus the middle point.


Third derivative $$\cal D_{t\, t\, t} = \frac{\cal D_{t\, t} f(t + h) - \cal D_{t\, t} f(t-h)}{2h}$$

$$= \frac{\frac{f(t+2h) -2 f(t+h) + f(t)}{2h} - \frac{f(t) - 2 f(t - h) + f(t - 2h)}{2h}}{2h}$$

$$= \frac{f(t + 2h) - 2\left({\large\strut} f(t+h) - f(t-h)\right) - f(t-2h)}{2 h^3}$$




Let's check if this is right ...

f(x + 2 * .h) - 2 * (f(x + .h) - f(x - .h)) - f(x - 2 * .h)/(2 *.h^3)
```{r}
f <- sin # Non-mosaiCalc notation
slice_plot(
  (f(t + 2*h) - 2*(f(t+h) - f(t-h)) - f(t - 2*h)) /
    (2*h^3) ~ t, 
  domain(t = -5:5), 
  h = 0.001
)
```

```{r}
numD(f(x) ~ x + x + x)
```




## If you have an analytic formula for a derivative, use it.

Example: $g(t) \equiv sin(t)$.

Algebraic fact : $\sin(a + b) = \sin(a)\cos(b) + \cos(a)\sin(b)$

Let's apply this to our definition of the derivative:

$$\frac{\sin(t + h) - \sin(t)}{h} = \frac{\sin(t)\cos(h) + \cos(t)\sin(h) - \sin(t)}{h}$$
$$ = \frac{(\cos{h} - 1)}{h} \sin(t) + \frac{\sin(h)}{h} \cos(t)$$

Example of a limit proof: [$\lim_{h \rightarrow 0} \frac{\sin(h)}{h} = 1$](https://www.youtube.com/watch?v=5xitzTutKqM)

## Close enough

### Sound

Sound is communicated to our ears by a function `pressure(t)`. What does "close enough for our purposes" mean here? 

[Recording of report about](https://archive.org/details/WWII_News_1940/1940-05-13_BBC_Bernard_Stubbs_On_The_BEF_Advance_Into_Belgium.mp3) the BEF entering Belgium (1940)

A pure sine wave

```{r}
time = (0:44100)/44100
hz = 440
audio::play(sin(hz*2*pi*time), rate = 44100)
```

For the purpose of representing audio to humans, we can cut out sounds that we can't perceive.

## Digital audio: Nyquist sampling theorem

```{r}
sinc <- makeFun(sin(t)/t ~ t)
slice_plot(sinc(t) ~ t, domain(t = -25:25), npts = 500)
```

### Training neural networks 




## Digital signal processing


```{r}
audio::play(Cello$y)

Sys.sleep(1)
Dcello <- (Cello$y[-1] - Cello$y[-nrow(Cello)])



audio::play(as.integer(20*Dcello))

```
