---
title: "Bond ratings ups and downs"
author: DTK
date: last-modified
format: live-html
desc: "Do bond ratings get higher?"
categories: [finance, linear algebra, ]
---

```{r include=FALSE}
library(dplyr)
```

The ratings of companies and governments by agencies such as Moody's Investor Services or  Standard and Poor's are used by investors to indicate the level of risk of loaning money (through, say, a bond).

The ratings take the form of letter grades as in @tbl-letter-grades-S-and-P.

::: {#tbl-letter-grades-S-and-P}
Grade | Financial strength description
-----:|:-------------------------
AAA | Extremely strong
AA | Very strong
A | Strong
BBB | Adequate
BB | Facing financial uncertainty
B | Vulnerable
CCC | Vulnerable and dependent on good business conditions to continue
CC | Highly vulnerable
C | Highly vulnerable to non-payment with low expectation of recovery
D | General default or breach
NR | Not rated

Rating grades and their interpretion for Standard and Poor's
:::

The letter grade is often used to set the interest rate for the loan, a combination of a "safe" interest rate (generally, the US Treasury Bill rate) and a premium to reflect risk. Low-rated companies have to pay a higher interest premium.

Bonds are often long-term affairs, spanning several years or even a decade or more. But a company's fortunes can change over the years; it's helpful to know, for example, the probability that a strongly-rated company will fall to a lower rating.

The transition matrix below was published by S&P in 2017 and reflects the changes in ratings over a three-year period. Prof. Haiyan Su (of Montclair State University) describes the use of the data in an "industrial mathematics" course in an upcoming AMS/MAA Classroom Resource Materials volume edited by Tim Chartier and Jason Douma, *Data Science: Plug-and-play Modules for the College Classroom*.

The `Transitions` data frame states the probability over the three year period of a company moving from one grade to another. The starting grade is given in the first column, and the different ending grades are the columns. 




```{r}
Transitions <- read.table(header = TRUE,  text = "
from  AAA   AA    A     BBB   BB    B     CCCC  D   NR 
AAA   .6548 .2209 .0235 .0032 .0019 .0008 .0011 .0013 .0924 
AA    .0121 .6614 .1853 .0206 .0035 .0022 .0003 .0012 .1133 
A     .0006 .0407 .6885 .1172 .0130 .0044 .0009 .0025 .1321
BBB   .0002 .0028 .0842 .6466 .0711 .0164 .0030 .0087 .1670 
BB    .0001 .0006 .0051 .1108 .4704 .1158 .0125 .0396 .2451 
B      0    .0003 .0021 .0078 .1023 .4146 .0467 .1257 .3005
CCCC   0     0    .0014 .0061 .0163 .1686 .1054 .4065 .2957"
)
```

```{r}
tmat <- Transitions |> 
  select(-from, -D, -NR) |>
  as.matrix() %>% t()
dim(tmat)
```

12-year transition simulation.
```{r}
tmat %*% tmat %*% tmat %*% tmat
```

```{r}
Factorization <- eigen(tmat)
Factorization
```


Task: Find the one-year transition matrix.
