::: {.panel-tabset}

```{r}
#| caption: For scratch work

```

It helps to distinguish between two settings for arguments in R functions.

Setting 1.
: A formal argument name. The name in the function definition is just for accounting purposes. The user doing a symbolic calculation can use any name in the tilde expression, so long as the tilde expression is internally consistent. For instance

```{r}
h <- makeFun(z^2 ~ z)
h
```


Setting 2. 
: A named argument. The name in the function is pre-specified and assigned a default value. For instance:

```{r}
g <- makeFun( z^2 ~ z, z = 2)
g # print it
```

Which of these commands will correctly create the anti-derivative of the given function.

a. `antiD(h(w) ~ w)`

b. `antiD(g(z) ~ z)`

c. `antiD(g(w) ~ w)`

d. `antiD(g(z = w) ~ w)`
 
e. `antiD(h(z = w) ~ w)`




Construct a function $f(y)$ in R. You can use a basic modeling function or whatever you like.

1. What kind of thing will be produced by `antiD(f(x) ~ x)`?

```{mcq}
#| label: index-accumulation-1FE
#| inline: true
1. A function of `y`.
2. A function of `x` [correct]
3. A numerical quantity.
4. The command is invalid. 
```

[Even though the function `f(y)` was constructed with `y` as the formal argument, you can use whatever argument name you like in the tilde expression.]{.shortanswer `r answer_style()` data-tooltip="Ans id: index-accum-6qK"}


2. What kind of thing will be produced by `antiD(f(y))`?

```{mcq}
#| label: index-accumulation-2FE
1. A function of `y`.
2. A function of `x`
3. A numerical quantity.
4. The command is invalid [correct]
```


:::



:::::
<!-- accumulation -->
