---
author: DTK
date: 2025-01-14
number-offset: 3
format: live-html
---

# 2025-01-14 class notes

```{r include=FALSE}
library(mosaicCalc)
```

## Where we are

Activities we didn't get to on Friday

::: {#lst-slice-plot}
```{webr-r}
#| caption: Making a graph of f(x).
#| context: interactive
#| persist: true
#| editor-code-line-numbers: 2
slice_plot(__tilde_exp__, 
           domain(x = -5:5))
```
:::

1. Plot out each of the pattern-book functions on a suitable domain. pnorm(), dnorm(), exp(), sin(), sqrt(), log(), 

2. Define a `recip()` function using `makeFun()`. 

3. Define an `identity()` function using `makeFun()`. (The identity function returns its input unaltered.)

4. Define a `power_law()` function using `makeFun()`. Example of a power-law function: $y^{2.4}$

## Spaces, graphs, and graphics

Composition of input and output spaces in `slice_plot()`. [link](https://www.mosaic-web.org/MOSAIC-Calculus/Textbook/Preliminaries/04-graphs-and-graphics.html#spaces-and-graphs-of-functions)

Mapping from (part of?) input space to output space.

### 2-d input spaces. 

- Can compose input space with output space to get a 3-dimensional space. Mathematicians like this because the surfaces are beautiful, especially when you can interact with them. 
    a. Example: [](https://www.mosaic-web.org/MOSAIC-Calculus/Textbook/Preliminaries/04-graphs-and-graphics.html#fig-first-surface){target="_blank"}
    b. The above example is quite complicated and relates to physics/engineering: a wave packet, and such.
    c. Of more general use in modeling are simpler functions that resemble terrain **locally**. The word "local" will be important. Calculus provides tools for translating between the local and the global.
    
```{r}
f <- doodle_fun(~ w & z, seed=101)
surface_plot(f(w, z) ~ w & z, domain(w=-5:5, z=-5:5))
```

We will tend to use "contour plots" instead. It's easier to make measurements on them, but you have to learn how.

```{r}
contour_plot(f(w, z) ~ w & z, domain(w=-5:5, z=-5:5), fill_alpha = 0.5)
```



[](https://www.mosaic-web.org/MOSAIC-Calculus/Textbook/Preliminaries/04-graphs-and-graphics.html#fig-contour-pre-slice)


