---
id: "boy-cut-sheet"
created: "Sat Apr 26 10:14:13 2025"
attribution: TBA
---

```{r include=FALSE, eval=!"devoirs" %in% (.packages())}
# For stand-alone compilation testing
library(devoirs)
library(mosaicCalc)
```

::: {#exr-boy-cut-sheet}
## Variance

The **variance** of a variable (in statistics) describes how much variability there is among the specimens. We're going to look at the variance of **vectors**.

Some easy computation will be needed. For instance, the variance of the vector $(1, 0, -1)^T$ is ...

```{r}
#| caption: variance calculations
var(vec(1,0,-1))
```

1. Calculate the variance of the vector $(5, 4, 3)$. Compare the result to the variance of $(1, 0, -1)^T$ and explain why this is happening. (Hint: It's not an error.)

`r devoirs_text("var-calc-1")`

The variance of a vector is closely related to the length of the vector. We'll show you a few attempts at making the approximating the relationship. Your job is to find the modification to the calculation that makes the vector-length method exactly the same as the variance.

```{r}
#| caption: vector length and variance
b <- vec(5,4,3)
veclen(b)^2 / nrow(b)
veclen(vec_mm(b))^2 / nrow(b)
```

2. What does `vec_mm()` do? And how can you change the denominator of the last line in the chunk so that the vector-length-style calculation corresponds *exactly* to the variance?

`r devoirs_text("var-calc-2")`

:::
 <!-- end of exr-boy-cut-sheet -->
