---
author: DTK
date: 2025-04-25
number-offset: 41
format: live-html
webr:
  channel-type: 'automatic'
  repos: ["https://dtkaplan.r-universe.dev"]
  packages: ['ggplot2', 'mosaicCalc', "LSTbook" ]
filters:
  - webr
---

{{< include ../../_extensions/r-wasm/live/_knitr.qmd >}}

```{r include=FALSE}
library(mosaicCalc)
library(devoirs)
library(dplyr)
library(LSTdata)
source("../../_software/software.R")
```
::: {.callout-tip collapse=true}
## Background software
```{webr-r}
#| autorun: true
#| context: output
{{< include ../../_software/software.R >}}
```
:::

[NYTimes article on birthrate](https://www.nytimes.com/2025/04/23/health/birth-rates-cdc.html?smid=nytcore-ios-share&referringSource=articleShare)


- Review `cumsum()` and `cumprod()`
- What other things amount to these operations: inflation, debt, mortgage
- `cumprod()` is (for positive numbers),

```{r}
Survive <- SSA2021F |>
  mutate(survival = 1 - mortality) |>
  mutate(total = exp(cumsum(log(survival))))
```

```{r}
mfun <- spliner(mortality ~ age, data = SSA2021F)

f <- D(mfun(age) ~ age + age)
slice_plot(f(age + 1) ~ age, domain(age = 1:120))
dmfun <- D(mfun(age) ~ age) 
ADmfun = antiD(log(1 - mfun(age)) ~ age)
slice_plot(exp(ADmfun(age)) ~ age, domain(age = 1:120))
```





```{r}
Survive |>
  gf_line(total ~ age) |> gf_point(total ~ age) 
```


```r
variable |> log() |> cumsum() |> exp()`
```

```r
SSA2021F |>
  mutate(survival = 1 - mortality) |> 
  mutate(total = survival |> log() |> exp())
```

antiD is like cumsum()

integrate is like sum()

Take care with traditional notation ...

- $\int g(t) dt$ means cumsum(): result is function
- $\int_0^{30} g(t) dt$ mean sum() the rows for ages 0 to 30. Result is a number.
- $\int_0^x g(t) dt$ means cumsum(), starting at zero, moving up to whatever is given for x. Result is a function.

Do calculation by splining, anti-differentiation, to get survival function.

Life expectancy at a given age.  

- For median death age, solve to find age that gives 50% of the starting mortality.
- For life expectancy, accumulate survival times age.

```{r}
Wartime <- Mortality_history |>
  filter(year == 1944)
mfun1944 <- spliner(mortality ~ age, data = Wartime)
f <-  antiD(log(1 - mfun(age)) ~ age)
slice_plot(exp(f(age)) ~ age, domain(age = 2:60)) |>
  slice_plot(exp(ADmfun(age))~age, color = "blue")
f
```

```{r}
g <- mosaic::rfun(~t, seed = 932)
slim <- makeFun(60 + g(t) ~ t)
slice_plot(slim(t) ~ t, domain(t = 0:3))
pos <- antiD(slim(t) ~ t)
slice_plot(pos(t) ~ t, domain(t = 0:3))
```

# 2025-04-25 class notes

```{webr-r}
#| caption: User console A
#| persist: true
```

```{webr-r}
#| caption: User console B
#| persist: true
```

```{webr-r}
#| caption: User console C
#| persist: true
```


