---
author: DTK
date: 2025-04-08
number-offset: 33
format: live-html
---



```{r include=FALSE}
library(mosaicCalc)
library(purrr)
library(LSTbook)
source("../../_software/software.R")
```



```{webr-r}
#| autorun: true
#| context: setup
{{< include ../../_software/software.R >}}
```


# 2025-04-08 class notes

::: {.callout-tip collapse=true}
## Console A
```{webr-r}
#| persist: true

```
:::

::: {.callout-tip collapse=true}
## Console B
```{webr-r}
#| persist: true

```
:::

::: {.callout-tip collapse=true}
## Console C
```{webr-r}
#| persist: true

```
:::


## From last week ...

We talked about the dot product, linear combinations, the idea of a subspace as defined by a set of vectors. 

We started to think about linear modeling (as in `model_train()`, which uses `lm()`) in terms of vectors.

Reminder of "case space" versus "variable space."


Today's goal is to give you a complete understanding of linear modeling in terms of vectors. 

## Matrices

A matrix is a set of vectors. All of them must share an embedding space, that is, they all must have the same number of components.

* Making vectors, say with `rvec()`
* Making matrices with `cbind()`
* Sometimes it will be nice to display a large matrix visually, one pixel per number.


Example: Matrix of random numbers

```{webr-r}
Mrand <- matrix(rnorm(40*30), nrow=40, ncol=30)
draw_matrix(Mrand)
```

## Grabbing components of matrices

Occasionally, we will need to look at just one (or a few) columns (or rows) in a matrix. We can use the **indexing syntax** for this, which involves square braces: `[ , ]`

Example: Matrix of small integers

```{r}
M <- values_mat(4, 3)
M
M[1, ]  # first row
M[ , 1] # first column
```

Now for an R oddity. The people who initially wrote S (the precursor to R) created an object called a "vector" that, unlike the linear algebra conception 


## Matrix/vector multiplication

Canonical form: m-by-n matrix times an n-dimensional vector gives an m-dimensional vector.

The result is a *linear combination of the vectors in the matrix*, with the "coefficients"/"scalars" coming from the vector. 

Not necessarily commutative. 

Why I was dodgy in writing the dot product using two column vectors. Better if the dot product is seen as a kind of matrix/vector multiplication, where the matrix consists simply as a row of scalars.

Use `values_mat()` to create the vectors and the matrix.

# Finding the coefficients using `lm()`

```{webr-r}
v <- values_mat(4, 1)
M <- values_mat(4, 4)
list(M=M, v=v) |>
  model_train(v ~ M - 1) |>
  conf_interval()
```

How does `model_train()` manage this? It calls a more basic function `lm()` (short for "linear model") which in turn calls `qr.solve()` which in turn calls a linear algebra package called "LINPACK" or a newer system "LAPACK" that's used by technical workers world-wide.

```{webr-r}
qr.solve(M, v)
```

## Q and R matrices

`qr.solve()` takes the matrix and re-writes it as a linear combination of two special matrices, one called Q and the other R. 

```{r}
myQ <- qr.Q(qr(M))
```


## Functions as vectors

Gaussians over a two second-window

```{r}
step = 2^(1/12)

Sounds <- tibble::tibble(
  time = (0:44100)/22050,
  g1 = pmin(1.0, dnorm(time, mean = 1, sd=0.5)),
  g2 = pmin(1.0, dnorm(time, mean = 0.25, sd=0.2)),
  g3 = pmin(1.0, dnorm(time, mean = .75, sd=0.2)),
  g4 = pmin(1.0, dnorm(time, mean = 1.25, sd=0.1)),
  one =    sin(step^0  * 2 * pi * 440 * time),
  two =    sin(step^1  * 2 * pi * 440 * time),
  three =  sin(step^2  * 2 * pi * 440 * time),
  four =   sin(step^3  * 2 * pi * 440 * time),
  five =   sin(step^4  * 2 * pi * 440 * time),
  six =    sin(step^5  * 2 * pi * 440 * time),
  seven =  sin(step^6  * 2 * pi * 440 * time),
  eight =  sin(step^7  * 2 * pi * 440 * time),
  nine =   sin(step^8  * 2 * pi * 440 * time),
  ten =    sin(step^9  * 2 * pi * 440 * time),
  eleven = sin(step^10 * 2 * pi * 440 * time),
  twelve = sin(step^11 * 2 * pi * 440 * time)
)  |>
  mutate(sum = g2*three + g2*five + g2*seven + 2*three^2)

audio::play(Sounds$sum / 10, rate = 22050)
```





## SVD and the dimension of subspaces

Create some vectors, then some linear combinations of them than look at the singular values.

```{webr-r}
u <- rbind(1,2,3, 6, -2)
v <- rbind(2,5,3, -4, 1)
w <- 3*u - 2*v
x <- w - u
y <- u + v - w
M <- cbind(u, v, w, x, y)
M |> svd() |> purrr::pluck("d")
```

```{webr-r}
draw_matrix(M)
M
```

```{webr-r}
M <- Sounds |> 
  select(one, two, three, four, five, 
         six, seven, eight, nine, ten, eleven, twelve) |>
  as.matrix(ncol = 12) 
M |> svd() |> purrr::pluck("d")
```


