---
id: "fir-bathe-jacket"
created: "Mon Jun  9 17:01:50 2025"
attribution: TBA
---

```{r include=FALSE, eval=!"devoirs" %in% (.packages())}
# For stand-alone compilation testing
library(devoirs)
library(mosaicCalc)
```

::: {#exr-fir-bathe-jacket}

Consider this dynamical system, whose flow is graphed below for $\mu = 1$:

$$\partial_t x = \mu\left(x - \frac{1}{3} x^3 - y\right)$$
$$ \partial_t y = \frac{1}{\mu} x$$


```{r}
#| warning: false
#| echo: false
mu <- 1
vdp <- makeODE(dx ~ mu*(x - x^3/3 -y) , dy ~ x/mu, mu = mu)
streamlines(vdp, domain(v=-2:2, x = -2.3:2.3), 
            npts = 25, alpha = 0.1, dt=.03) |>
  contour_plot(
    mu*(1-x^2)*v - x ~ v & x, 
    domain(v=-1.4:1.4, x = -1.4:1.4), mu=mu,
    filled = FALSE, labels = FALSE, contours_at = 0) |>
  gf_refine(coord_fixed())
```  
a. There is only one fixed point in the system. Where is it?

`r vspace(2)`

b. Part of one of the nullclines is shown in the figure. Which is it?

`r vspace(2)`

c. The flow has places where nearby trajectories (i) get pushed together and places where nearby trajectories are (ii) pulled apart. Circle a small region of the state space where (i) applies and another where (ii) applies.

d. Starting from near (not at!) the fixed point draw a trajectory for the flow.

e. The Jacobian matrix corresponds to a linearization of the flow near the fixed point. It looks like this:

$$\left(\begin{array}{r}
\partial_{xt} x & \partial_{yt} x\\
\partial_{xt} y & \partial_{yt} y
\end{array}\right)\ \text{evaluated at the fixed point}$$

Use the dynamical functions to calculate the Jacobian and write it here:

`r vspace(5)`

f. The eigenvalues of the Jacobian are $$\lambda_{1,2} = \frac{ \mu\pm\sqrt{\mu^2 - 4}}{2}$$
    i. Based on the eigenvalues, when $\mu = 1$ what is the stability of the fixed point? Use terms like these: Stable cycle, stable fixed point, unstable cycle, unstable fixed point, 
    ii. Think about increasing $\mu$. At what value of $\mu$ does the stability undergo a *qualitative change* in stability from that in (i)?
    iii. As $\mu$ increases above 2, what is the stability near the fixed point?

:::
 <!-- end of exr-fir-bathe-jacket -->
