---
id: "birch-ring-pen2"
created: "Mon Jun  2 16:11:54 2025"
attribution: TBA
---

```{r include=FALSE, eval=!"devoirs" %in% (.packages())}
# For stand-alone compilation testing
library(devoirs)
library(mosaicCalc)
```

::: {#exr-birch-ring-pen-2}

@fig-f-g-203 shows a flow field and some nullclines.

::: {#fig-f-g-203}

```{r}
#| echo: false
f <- doodle_fun(~ x & y, seed = 203)
g <- doodle_fun(~ x & y, seed = 679)
dom = domain(x = -3:4, y = -2.5:8)
streamlines(dx ~ f(x,y), dy ~ g(x,y), dom, dt = 0.025) |>
contour_plot(f(x,y) ~ x & y, dom, contours_at =  0, contour_color = "magenta") |>
  contour_plot(g(x,y) ~ x & y, dom, contours_at =  0, contour_color = "darkgreen") |>
  gf_theme(theme_minimal())
  
```

:::

Are the flow and the nullclines shown in @#fig-f-g-203 plausibly from the same dynamical system? Explain what you see in the graph that leads you to your conclusion.

`r devoirs_text("f-g-203")`

<!-- Opening answer: bp2-7OX -->
::: {.callout-tip collapse=true `r answer_style()`}
## Answer

They match.

1. The flow shows no sign of converging to or diverging from a fixed point, so no fixed point is evident. This is consistent with their being no intersection of the two nullclines.

2. The flow crossing one of the nullclines is entirely vertical, the flow crossing the other is entirely horizontal.


[..id..]{data-tooltip="Ans id: bp2-7OX"}
:::
<!-- closing answer bp2-7OX -->




:::
 <!-- end of exr-birch-ring-pen2 -->
