---
author: DTK
date: 2025-01-28
number-offset: 9
format: live-html
---

# 2025-01-28 class notes

```{r include=FALSE}
library(mosaicCalc)
```

[picked up from Friday's class]


## Chapter 11: Fitting features

Input- and output-scaling heuristics for a handful of pattern book functions 

Note that the output scaling is a form of LINEAR COMBINATION.

- Exponential [We did this one on Friday 24 Jan.]

```{webr-r}
t0 <- sample(1:20, 1)
halflife <- exp(runif(1, -15, 15))
span <- sort(runif(2, t0 - 5*halflife, t0 + 5*halflife ))
f <- makeFun(A*exp((t-t0)/halflife) + B ~ t, 
             A = sample(-10:10,1), 
             B = sample(-10:10, 1))
slice_plot(f(t) ~ t, 
           domain(t = c(span)),
           npts = 500
           )
```   

- Sin(). There are four parameters. 
     -The output and output scaling is the same as for other functions, but a different convention is used for the input scaling expression: $$A \sin(2\pi (t-t_0)/P) + B .$$
        - $P$ is the **period**: peak-to-peak in the domain.
        - $B$ is still called the "baseline" but it is not at the bottom of the function range. 
       - $A$ is the "amplitude". It corresponds to **half** the peak-to-trough range.
       - $t_0$ is the "left-shift," the location in the domain of an upward-sloping baseline crossing. 
       - Sometimes people write things differently: 
           i. $A \sin(2\pi (t-t_0)/P) + B$
           ii. $A \sin(2\pi t/P + \theta) + B$
           #. $A \sin(\omega t + \phi) + B$
           #. $A \sin(\omega t) + B \cos(\omega t) + C$
        - The first three of the above should be recognizably equivalent with simple algebraic re-arrangement. (iv) is more subtle and results from a trigonometric *identity*. 
        - We'll use (i), and you're expected to be able to estimate by eye the four parameters.
        - We could also write $A \cos(2 \pi (t - t_0)/P) + B$, but $t_0$ would be the location of a peak.
       
    
```{webr-r}
period <- sample(1:20, 1)
span <- sort(runif(2, -3*period, 3*period ))
slice_plot(A*sin(2*pi*(t-t0)/P) + B ~ t, 
           domain(t = c(span)),
           A = sample(-10:10, 1), 
           B = sample(-10:10, 1),
           P = period, 
           t0 = sample(seq(-pi, pi, length=10), 1),
           npts = 500
           )
```           



- Gaussian & sigmoid
    - Output scaling is the same as always: $A$ and $B$.
        - for $A > 0$, $B$, "baseline," is the value at the left horizontal asymptote.
        - $A$ is more difficult for `dnorm()`
            - `pnorm()`: easy. Vertical distance between lower and upper horizontal asymptote. 
            - `dnorm()` the meaning of $A$ depends on `sd`. If you must know (you don't at this point), 
    - Input scaling is different, written in terms of two *new* parameters: `mean` and `sd`.
        - `mean` the location in the domain of the peak (`dnorm()`) or the half-way point (`pnorm()`)
        - `sd`, the "standard deviation," is 
            - `dnorm()`: *roughly* the half-width (61%) at half-height 
            - `pnorm()`: *roughly* the width between the half-way point and the input producing an output at three-quarters (80%) of the way to the top.
    

```{r}
mean <- sample(1:20, 1)
sd <- runif(1, 0.5, 5)
fun <- ifelse(rnorm(1) > .25, pnorm, dnorm)
span <- runif(1, -mean - 3*sd, mean - sd)
span[2] <- runif(1, mean + sd, mean + 3*sd )
f <- makeFun(A*fun(t, mean=mean, sd==sd) + B ~ t, 
             A = sample(setdiff(-10:10, 0),1), 
             B = sample(-10:10, 1))
slice_plot(f(t) ~ t, 
           domain(t = c(span)),
           npts = 500
           )
```           


## Chapter 10: Functions with multiple inputs

Making them by ...

- Linear combination
- Multiplication
- Composition (?)
- Piecewise (?)

Common forms from science generally:

- Example: Breaking distance of car as function of speed and road conditions (slip)

- Interaction. 
    - Example, rabbits and foxes. Change in rabbit population as function of rabbit population and fox population.
    - Example, rate of reaction between two re-agents $\alpha\ [A][B]$
    - Example (maybe): demand for one good as a function of price of that good and price of a substitute.
    - We are going to see this particularly when we look at low-order polynomials in multiple variables.
    
    



