---
author: DTK
date: 2025-05-29
number-offset: 52
format: live-html
webr:
  channel-type: 'automatic'
  repos: ["https://dtkaplan.r-universe.dev"]
  packages: ['ggplot2', 'mosaicCalc', "LSTbook" ]
filters:
  - webr
---

{{< include ../../_extensions/r-wasm/live/_knitr.qmd >}}

```{r include=FALSE}
library(mosaicCalc)
library(devoirs)
source("../../_software/software.R")
```
::: {.callout-tip collapse=true}
## Background software
```{webr-r}
#| autorun: true
#| context: output
{{< include ../../_software/software.R >}}
```
:::

## Meanwhile, back on the ranch ...
```{r}
consumption <- 
  makeFun((beta*v^2/(v0^2 + v^2))~ v, 
          beta=0.1, v0=1)
slice_plot(consumption(v) ~ v, bounds(v=0:10)) %>%
  gf_labs(y="Consumption (tons/day)", x="v: available biomass (tons)")

```


Find the bifurcation point in $H$.

```{r}
#| message: false
Grazing <- makeODE(dV ~ r*V*(1 - V/K) - H*consumption(V), 
                    H=36, r = 2,  K=10)
TS <- integrateODE(Grazing, domain(t=0:20), V=.4)
traj_plot(TS$V(t) ~ t, domain(t=0:20)) # or slice_plot()
```
 a was 0.0026
```{r}
SIR <- makeODE(dS~ -a*S*I, dI ~ a*S*I - b*I, a=0.001, b=0.5, S=762, I=10)

Traj <- integrateODE(SIR, domain(t=0:50), I = 100)
flow_field(SIR, domain(S=0:800, I = 0:300), npts = 15)
  traj_plot(Traj$I(t) ~ Traj$S(t), domain(t=0:50))
  traj_plot(Traj$S(t) ~ t, domain(t=0:50))
```








## Predator-Prey 

$$\frac{dx}{dt} = \alpha x - \beta x y,\\
\frac{dy}{dt} = - \gamma y + \delta x y , $$ 

alpha=0.1 "Reproduction rate of prey";
beta=0.02 "Mortality rate of predator per prey";
gamma=0.4 "Mortality rate of predator";
delta=0.02 "Reproduction rate of predator per prey";
x0=10 "Start value of prey population";
y0=10 "Start value of predator population"


## Richer State spaces

Functions of plucked string





## Waves in space
                      
[Maxwell's equations](https://en.wikipedia.org/wiki/Maxwell%27s_equations) (1861)
: $$\begin{align}
\nabla \cdot \mathbf{E} \,\,\, &= \frac{\rho}{\varepsilon_0} \\
\nabla \cdot \mathbf{B} \,\,\, &= 0 \\
\nabla \times \mathbf{E} &= -\frac{\partial \mathbf{B}}{\partial t} \\
\nabla \times \mathbf{B} &= \mu_0 \left(\mathbf{J} + \varepsilon_0  \frac{\partial \mathbf{E}}{\partial t} \right)
\end{align}$$

In the absence of charge ($\rho$) and current ($\mathbf J$), construct $\partial_{tt}{\mathbf E}$ and see what that tells you.


# 2025-05-29 class notes

```{webr-r}
#| caption: User console A
#| persist: true
```

```{webr-r}
#| caption: User console B
#| persist: true
```

```{webr-r}
#| caption: User console C
#| persist: true
```


