---
author: DTK
date: 2025-04-10
number-offset: 34
format: live-html
webr:
  channel-type: 'automatic'
  repos: ["https://dtkaplan.r-universe.dev"]
  packages: ['ggplot2', 'mosaicCalc', "LSTbook" ]
filters:
  - live
---

{{< include ../../_extensions/r-wasm/live/_knitr.qmd >}}

```{r include=FALSE}
library(mosaicCalc)
library(purrr)
library(LSTbook)
library(devoirs)
{{< include ../../_software/software.R >}}
```



# 2025-04-10 class notes

```{webr-r}
#| caption: User console A
#| persist: true
```

```{webr-r}
#| caption: User console B
#| persist: true
```

```{webr-r}
#| caption: User console C
#| persist: true
```


```{r}
#| results: asis
cat(htmltools::HTML("The square root is $\\sqrt{x}$"))
```

<form id="radioForm">
  <label><input type="radio" id="option1" name="options" value="value1"> $\sqrt{x^2}$</label><br>
  <label><input type="radio" id="option2" name="options" value="value2"> If I really, really, really, wanted to say something about this then I would type in a very long answer such as $\int_0^\infty \frac{dx}{x}$</label><br>
  <label><input type="radio" id="option3" name="options" value="value3"> Option 3</label><br>
</form>

 


And here is the second question ...

<form id="secondForm">
  <label><input type="radio" id="option4" name="options2" value="value4"> Option A</label><br>
  <label><input type="radio" id="option5" name="options2" value="value5"> Option B</label><br>
  <label><input type="radio" id="option6" name="options2" value="value6"> Option C</label><br>
</form>

<script type='text/javascript'>
    document.addEventListener('DOMContentLoaded', function() {
      const form = document.getElementById('radioForm');
      const radioButtons = form.querySelectorAll('input[type="radio"]');
    
      // Load saved value on page load
      const savedValue = localStorage.getItem('selectedOption');
      if (savedValue) {
        radioButtons.forEach(radio => {
          if (radio.value === savedValue) {
            radio.checked = true;
          }
        });
      }
    
      // Save value on change
      form.addEventListener('change', function(event) {
        if (event.target.type === 'radio') {
          localStorage.setItem('selectedOption', event.target.value);
        }
      });
    });
</script>

<script type='text/javascript'>
    document.addEventListener('DOMContentLoaded', function() {
      console.log("Adding event listener.");
      const form = document.getElementById('secondForm');
      const radioButtons = form.querySelectorAll('input[type="radio"]');
    
      // Load saved value on page load
      const savedValue = localStorage.getItem('secondFormOption');
      if (savedValue) {
        console.log("Restoring radio button values.");
        radioButtons.forEach(radio => {
          if (radio.value === savedValue) {
            radio.checked = true;
          }
        });
      }
    
      // Save value on change
      form.addEventListener('change', function(event) {
        if (event.target.type === 'radio') {
          localStorage.setItem('secondFormOption', event.target.value);
        }
      });
    });
</script>

`r devoirs_text("try-persistent")`

<script type="text/javascript">

      //save entered info
      document.getElementById("try-persistent").addEventListener("input", function () {
        var todo = document.getElementById("try-persistent").value;
        localStorage.setItem("try-persistent", todo);
      }, false);

      //on page load, read in the old value
      const savedValue = localStorage.getItem('try-persistent');
      if (savedValue) {
        document.getElementById("try-persistent").value = savedValue
      }
      

    </script>

## From last week ...

